---
title: "Statistical Analyses"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(modelr)
library(mgcv)

options(scipen = 50)
knitr::opts_chunk$set(echo = TRUE)
```
<p/>
<p/>
# Regression

Here, we will conduct linear regression analyses to understand what factors predict time spent on leisure activities (hours/day).
<p/>
<p/>

### Univariate Analysis

First, we will explore the demographics of the respondents included in our database and distribution of the predictors. We will check to make sure that age can be used as a continuous variable by plotting age against time spent on leisure activities. From the plot below, we don't see any clear distinctions indicating a need for categorization of age, so we will leave it as a continuous variable, as there appears to be monotonic and linear relationship.

<<<<<<< HEAD
```{r univariate, message=FALSE, show_col_types = FALSE, warning = FALSE}
=======
```{r demographic table, message=FALSE, show_col_types = FALSE, warning = FALSE}
>>>>>>> ca40ac7a5cc5ee771d8265d1ac7ab252111f1dee
cps_summary = 
  read_csv("./data/summary_household_category.csv")

cps_summary_leisure = cps_summary %>% 
  filter(category == "leisure")

library(table1)
table1(~ race + age + sex + labor_force_status + region|year, data = cps_summary)
```

Respondents' demographic characteristics stayed the same from 2019 to 2020. This means that the survey sample did not change significantly. It is reasonable to compare how time spent varied pre and post pandemic using ATUS data. 

```{r univariate, message=FALSE, show_col_types = FALSE, warning = FALSE}
ggplot(data = cps_summary_leisure, aes(x = age, y = category_sum_hour)) + geom_point() +
  labs(x = "Age (yrs)",
       y = "Time Spent on Leisure Activities (hrs/day)",
       title = "Relationship between Age and Leisure Time")

```

### Linear Models

Next, we want to explore whether there are differences in time spent on leisure activities between 2019 and 2020, as we have hypothesized that in 2020, after the pandemic hit and the US went into lockdown, people spent more at home engaging in new activities and hobbies.

```{r year_model, message=FALSE, show_col_types = FALSE, warning = FALSE}
cps_summary_leisure$year =
  as_factor(case_when(
    cps_summary_leisure$year == 2019 ~ "Pre-Pandemic Onset 2019",
    cps_summary_leisure$year == 2020 ~ "Post-Pandemic Onset 2020",
    TRUE ~ "unknown"
  ))

model_year = lm(category_sum_hour ~ year, data = cps_summary_leisure)
model_year %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  mutate(term = str_replace(term, "^year", "Year: ")) %>% 
  knitr::kable(digits = 3)
```

There is a significant difference between 2019 and 2020 in terms of leisure time spent (p-value <0.05). We expect that leisure time spent in 2020 (post-pandemic onset) is 0.2 hours longer per day than leisure time spent in 2019, on average. This aligns with our hypothesis that after the pandemic, people started spending additional time on leisure activities.


Now, we would like to fit a few models that we hypothesize may predict time spent on leisure activities based on literature and own experiences.

**(Model 1)**: Labor force status
We believe that one of the strongest predictors of time spent on leisure activities is whether someone is currently employed or not. The categories of labor force status are employed but absent from work (reference group), retired, employed at work, unemployed and looking, and unemployed on layoff.

**(Model 2)**: Demographic characteristics: race, age, and sex
Race, age, and sex, are three common demographic variables that we would expend to see trends with for time spent on leisure activity.

**(Model 3)**: Demographic characteristics: race, age, and sex, and their interactions 
This model builds on model 2 by exploring whether there are any additive or multiplicative interactions present between the 3 demographic variables.

```{r model1, message = FALSE, warning = FALSE}
model1 = 
  lm(category_sum_hour ~ labor_force_status, data = cps_summary_leisure)

model1 %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>% 
  mutate(term = str_replace(term, "^labor_force_status", "Employment Status: ")) %>% 
  knitr::kable(digits = 3, caption = "Model 1 Parameter Estimates")
```

From the table above, we can see that in Model 1, employment status is a very strong predictor of time spent on leisure activities. Those who are employed and at work are expected to spend 1.4 fewer hours per day on leisure activities per day in comparison to those who are employed but absent from work, on average. Additionally, those who are retired are expected to spend 1.5 more hours per day on leisure activities per day in comparison to those who are employed but absent from work. 

```{r model2, message = FALSE, warning = FALSE}
model2 = 
  lm(category_sum_hour ~ age + race + sex, data = cps_summary_leisure)

model2 %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>%
  mutate(
    term = str_replace(term, "^race", "Race: "),
    term = str_replace(term, "sexMale", "Sex: Male"),
    term = str_replace(term, "age", "Age")
    ) %>% 
  knitr::kable(digits = 3, caption = "Model 2 Parameter Estimates")

```

The 3 main effect predictors in Model 2 are all significant (p-value < 0.05). Specifically, for every 1 year increase in age, we expect people to spend 0.06 hours more on leisure activities per day on average, adjusting for race and sex. We expect males to spend 0.8 hours more on leisure activities per day on average, adjusting for race and age. Finally, those of Hawaiian/Pacific Islanders descent are expected to spend 1.02 hours more per day on leisure activities when compared to people of mixed race, while those who are Asian spend the least time on leisure activities on average, adjusting for sex and age. 

```{r model3, message = FALSE, warning = FALSE}
model3 = 
  lm(
    category_sum_hour ~ age + race + sex + age*race + age*sex + race*sex + age*race*sex,
    data = cps_summary_leisure)

model3 %>% 
  broom::tidy() %>% 
  select(term, estimate, p.value) %>%
  mutate(
    term = str_replace(term, "^race", "Race: "),
    term = str_replace(term, "sexMale", "Sex: Male"),
    term = str_replace(term, "age", "Age"),
    term = str_replace(term, "Age:race", "Age * Race:"),
    term = str_replace(term, "Age:Sex", "Age * Sex"),
    term = str_replace(term, "Race: American Indian:Sex: Male", "Race:American-Indian * Sex:Male"),
    term = str_replace(term, "Race: Asian:Sex: Male", "Race:Asian * Sex:Male"),
    term = str_replace(term, "Race: White:Sex: Male", "Race:White * Sex:Male"),
    term = str_replace(term, "Race: Black:Sex: Male", "Race:Black * Sex:Male"),
    term = str_replace(term, "Race: Hawaiian/Pacific Islander:Sex: Male", "Race:Hawaiian/Pacific Islander * Sex:Male"),
    term = str_replace(term, "Age * Race:American Indian:Sex: Male", "Age * Race:American Indian * Sex:Male"),
    term = str_replace(term, "Age * Race:Asian:Sex: Male", "Age * Race:Asian * Sex:Male"),
    term = str_replace(term, "Age * Race:Black:Sex: Male", "Age * Race:Black * Sex:Male"),
    term = str_replace(term, "Age * Race:White:Sex: Male", "Age * Race:White * Sex:Male"),
    term = str_replace(term, "Age * Race:Hawaiian/Pacific Islander:Sex: Male", "Age * Race:Hawaiian/Pacific Islander * Sex:Male"),
    ) %>% 
  knitr::kable(digits = 3, caption = "Model 3 Parameter Estimates")
```

Upon including the interaction terms, race is no longer significant, as are none of the interaction terms (p-value > 0.05).
<p/>
<p/>
### Diagnostics

Before we proceed, it is also important to look at residuals and fitted values in our models. 
```{r residuals_plots, message = FALSE, warning = FALSE}
cps_summary_leisure %>% 
  modelr::add_residuals(model1) %>% modelr::add_predictions(model1) %>% 
  ggplot(aes(x = resid, y = pred)) + geom_point(alpha = 0.4) +
  labs(x = "Predicted Leisure Time Spent (hr)",
       y = "Residuals",
       title = "Model 1: Relationship between Predicted Leisure Time and Residuals")

cps_summary_leisure %>% 
  modelr::add_residuals(model2) %>% modelr::add_predictions(model2) %>% 
  ggplot(aes(x = resid, y = pred)) + geom_point(alpha = 0.4) +
  labs(x = "Predicted Leisure Time Spent (hr)",
       y = "Residuals",
       title = "Model 2: Relationship between Predicted Leisure Time and Residuals")

cps_summary_leisure %>% 
  modelr::add_residuals(model3) %>% modelr::add_predictions(model3) %>% 
  ggplot(aes(x = resid, y = pred)) + geom_point(alpha = 0.4) +
  labs(x = "Predicted Leisure Time Spent (hr)",
       y = "Residuals",
       title = "Model 3: Relationship between Predicted Leisure Time and Residuals")
```

There are some outliers that should be noted and addressed in future analyses. Additionally, we would hope to see a random distribution of residuals across the predicted leisure time values, however there appears to be clear patterns and we can see skewness in the 3 model. This indicates that our models can be improved, but for the purposes of this project, we will continue with our existing model comparisons.
<p/>
<p/>
### Model Comparison

In order to perform cross validation, we will first create the crossv_mc tibble with the train and test data.
```{r crossmv_mc, message = FALSE, warning = FALSE}
cv_df = crossv_mc(cps_summary_leisure, 100) %>% 
  mutate(
    train = map(train, as_tibble),
    test = map(test, as_tibble)
  )
```

Next, we will fit the models and get the root mean squared errors.
```{r fit, message = FALSE, warning = FALSE}
fitted_df = cv_df %>%
  mutate(
    model1 = map(train, 
                 ~lm(category_sum_hour ~ labor_force_status, 
                     data =.x)),
    model2 = map(train,
                  ~lm(category_sum_hour ~ age + race + sex, 
                     data = .x)),
    model3 = map(train,
                  ~lm(category_sum_hour ~ 
                      age + race + sex + age*race + age*sex + race*sex + age*race*sex,
                      data = .x))) %>% 
  mutate(
    rmse_model1 = map2_dbl(model1, test, ~rmse(model = .x, data = .y)),
    rmse_model2  = map2_dbl(model2, test, ~rmse(model = .x, data = .y)),
    rmse_model3  = map2_dbl(model3, test, ~rmse(model = .x, data = .y))
)
```

Finally, we will plot the distribution of RMSE values to compare each candidate model.
```{r rsme_plot, message = FALSE, warning = FALSE}
fitted_df %>%
  select(starts_with("rmse")) %>% 
  pivot_longer(
    everything(),
    names_to = "model", 
    values_to = "rmse",
    names_prefix = "rmse_") %>% 
  mutate(model = fct_inorder(model)) %>% 
  ggplot(aes(x = model, y = rmse, color = model)) + geom_violin(alpha = 0.5) +
  labs(
    x = "Regression Model",
    y = "RSME",
    title = "Leisure Time Spent: Regression Models and RSME values"
  )
  
```

From the violin plot, we can see that Model 1, the model with employment status as the predictor, has the lowest RMSE average, which means it is the best model fit of the 3 that we tested. The interaction terms did not seem to make much of a difference in terms of RSME, as indicated by similar Model 2 and Model 3 plots.
